---
title: "In-class Ex 5: Visualising Text"
author: "Liang Xiuhao Rydia"
date: "May 11, 2024"
date-modified: "last-modified"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
---
## Visualising and Analysing Text Data

## 1.0 Getting Started
```{r}
pacman::p_load(tidyverse,readtext,
               quanteda,tidytext,
               jsonlite,igraph)
```

-   Read about [`quanteda`](https://quanteda.io/)

### 1.1 Loading the all the articles into a single file 

Using `readtext()` to convert many articles into a data table, consisting of article name in first column and article:
```{r}
text_data <- readtext(paste0("data/articles",
                             "/*"))
```


```{r}
usenet_words <- text_data %>%
  unnest_tokens(word, text) %>%
  filter(str_detect(word, "[a-z']$"),
         !word %in% stop_words$word)
```


```{r}
usenet_words |> 
  count (word, sort = TRUE)
```

```{r}
text_data_splitted <- text_data |> 
  separate_wider_delim("doc_id",
                       delim = "__0__",
                       names = c("X","Y"),
                       too_few = "align_end")
```


## 1.2 Using `jsonlite` package 
- easy to glean the insight from the data through looking at the details of the data. - see time series difference

```{r}
mc1_data <- fromJSON("data/mc1.json")
```


```{r}
mc2_data <- fromJSON("data/mc2
                     .json")
```

For creating graph model:
- [igraph](https://igraph.org/)







